---
title: "Getting started with typethis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with typethis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`typethis` adds runtime type safety to R. This vignette is a 10-minute
tour of the four things you'll reach for most: type checks, typed
functions, typed models, and composable type specs.

```{r setup}
library(typethis)
```

## 1. Type checks

The lowest-level entry points test, assert, and validate types. Use them
at function boundaries, when parsing inputs, or anywhere a quick sanity
check is enough.

```{r}
is_type(5, "numeric")
is_type("hello", "numeric")

# assert_type() throws on mismatch
assert_type(5, "numeric", "x")

# validate_type() returns a list instead of throwing
validate_type("hello", "numeric", "x")
```

`is_type()` accepts any of:

- a builtin name — `"numeric"`, `"integer"`, `"character"`, `"logical"`,
  `"list"`, `"data.frame"`, `"matrix"`, `"factor"`, `"date"`, `"posixct"`,
  `"function"`, `"environment"`;
- a predicate function `function(value) -> logical`;
- a [type spec](type-specs.html) such as `t_union("integer", "character")`.

For safe coercion, use `coerce_type()`:

```{r}
coerce_type("123", "numeric")
coerce_type(c(1, 0, 1), "logical")
```

## 2. Typed functions

`typed_function()` wraps a function so each call is validated.

```{r}
greet <- typed_function(
  function(name, times = 1L) paste(rep(name, times), collapse = " "),
  arg_specs = list(name = "character", times = "integer"),
  return_spec = "character"
)

greet("hi", times = 3L)
```

A bad call fails before the body runs:

```{r, error = TRUE}
greet("hi", times = "three")
```

`typed_function()` supports every R calling convention — positional,
named, mixed, reordered, and `...` passthrough — and respects argument
defaults. Set `coerce = TRUE` to convert mismatches before checking, or
`validate = FALSE` for hot paths where you've already validated upstream.

For the full feature set see `?typed_function` and the
[validators-and-models](validators-and-models.html) vignette.

### Retrofitting existing functions

When you want to add type checks to a function you already have,
`as_typed()` removes the boilerplate of restating every argument. Specs
for arguments with literal atomic defaults are inferred automatically;
only arguments you want to override appear in the call.

```{r}
add <- as_typed(function(x = 0L, y = 0L) x + y, .return = "integer")
add(2L, 3L)
```

```{r, error = TRUE}
add("two", 3L)
```

Pass overrides by name. Inference fills in the rest:

```{r}
greet <- as_typed(
  function(name = "world", times = 1L) {
    paste(rep(name, times), collapse = " ")
  },
  name = t_vector_of("character", exact_length = 1L)
)
greet("hi", times = 3L)
```

Pass `name = NULL` to opt one argument out of validation, or
`.infer = FALSE` to disable inference entirely. See `?as_typed` and
`?infer_specs` for the full inference rules.

For an existing typed function, `types()` is a symmetric replacement-form
accessor over `as_typed()`:

```{r}
greet_clone <- function(name = "world", times = 1L) {
  paste(rep(name, times), collapse = " ")
}
types(greet_clone) <- types(greet)
is_typed(greet_clone)

# NULL un-types
types(greet_clone) <- NULL
is_typed(greet_clone)
```

To retrofit a whole environment in one call, `as_typed_env()` walks the
environment, applies `as_typed()` to every function it finds, and writes
the typed versions back. Per-function overrides flow through `.specs`:

```{r}
e <- new.env()
e$add <- function(x = 0L, y = 0L) x + y
e$greet <- function(name = "world") paste0("hi ", name)

as_typed_env(e, .specs = list(
  add = list(.return = "integer")
))
is_typed(e$add)
attr(e$add, "return_spec")
```

## 3. Typed models

Use `define_model()` to describe a record type — a configuration object,
an API request body, a domain entity. It generates a `new_<Class>()`
constructor and an `update_<Class>()` updater in the calling environment.

```{r}
define_model("User", fields = list(
  name = field("character"),
  age = field("integer",
    validator = numeric_range(0, 120),
    default = 0L
  ),
  email = field("character",
    validator = string_pattern("^[^@]+@[^@]+\\.[^@]+$")
  )
))

ada <- new_User(name = "Ada Lovelace", email = "ada@example.com")
ada
```

Validation runs at construction time:

```{r, error = TRUE}
new_User(name = "Ada", age = 200L, email = "ada@example.com")
```

`update_User()` returns a fresh instance with the named fields replaced
and revalidated:

```{r}
ada <- update_User(ada, age = 36L)
ada$age
```

Models nest. Reference one model from another by class name:

```{r}
define_model("Address", fields = list(
  street = field("character"),
  city   = field("character")
))

define_model("Person", fields = list(
  name    = field("character"),
  address = field("Address")
))

p <- new_Person(
  name    = "Ada",
  address = new_Address(street = "Newstead Abbey", city = "Nottingham")
)
p$address$city
```

For defaults, nullability, strict mode, and validators, see the
[validators-and-models](validators-and-models.html) vignette.

## 4. Composable type specs

The `t_*()` family builds richer specifications that work everywhere a
plain type name does (`is_type()`, `assert_type()`, `field()`,
`typed_function()`, `to_json_schema()`).

```{r}
id <- t_union("integer", "character")
is_type(1L, id)
is_type("u-42", id)

maybe_int <- t_nullable("integer")
is_type(NULL, maybe_int)

tags <- t_list_of("character", min_length = 1L)
is_type(list("alpha", "beta"), tags)

role <- t_enum(c("admin", "user", "guest"))
is_type("admin", role)
```

Specs compose:

```{r}
mixed <- t_list_of(t_union("integer", "character"))
is_type(list(1L, "two", 3L), mixed)
```

Use any spec inside a `field()` or as `typed_function()`'s `arg_specs`
entry. See the [type-specs](type-specs.html) vignette for the full
reference.

## Where to go next

- [`validators-and-models`](validators-and-models.html) — built-in
  validators, defaults, nullability, strict mode, the `field()` metadata.
- [`type-specs`](type-specs.html) — every `t_*()` constructor and how
  they compose.
- [`interop`](interop.html) — JSON Schema, Open Data Contract Standard,
  and OpenAPI 3.1 export and import.

For function reference, `?typethis` is a topic-grouped index and each
function's help page has a "See also" section linking to its family.
