---
title: "Interop: JSON Schema, ODCS, and OpenAPI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interop: JSON Schema, ODCS, and OpenAPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(typethis)
```

`typethis` ships three bridges so the same type definitions can travel
out to non-R systems and (for ODCS and OpenAPI) come back in:

- **JSON Schema** (Draft 2020-12) via [`to_json_schema()`].
- **Open Data Contract Standard v3** via [`to_datacontract()`] and
  [`from_datacontract()`].
- **OpenAPI 3.1** via [`to_openapi()`] and [`from_openapi()`].

All three share the same machinery — JSON Schema is the canonical
representation, and the ODCS and OpenAPI exporters lift its fragments
into the appropriate envelope.

The optional dependencies (`jsonlite`, `yaml`) live in `Suggests`. The
examples below guard their use with `requireNamespace()`.

## JSON Schema

`to_json_schema()` turns typed models, type specs, validators, and
[`field()`] definitions into a Draft 2020-12 JSON Schema fragment as a
plain R list.

```{r}
define_model("Person", fields = list(
  name = field("character", nullable = FALSE),
  age  = field("integer", validator = numeric_range(0, 120)),
  role = field(t_enum(c("admin", "user")), default = "user")
))

schema <- to_json_schema("Person")
str(schema, max.level = 2)
```

```{r, eval = requireNamespace("jsonlite", quietly = TRUE)}
cat(jsonlite::toJSON(schema, auto_unbox = TRUE, pretty = TRUE))
```

Key behaviours:

- Built-in validator constraints (`numeric_range`, `string_length`,
  `string_pattern`, `vector_length`, `enum_validator`) become native
  JSON Schema keywords (`minimum`, `maxLength`, `pattern`, `minItems`,
  `enum`).
- Nested models become `$ref` entries in `$defs`. Cyclic references are
  handled via a stub-then-fill protocol.
- Constructs without a canonical mapping (data frames, factors, custom
  predicate functions) emit `x-typethis-*` extension keys so they
  round-trip through typethis-aware tooling.

Other accepted inputs include type specs (`to_json_schema(t_union(...))`),
validator closures (`to_json_schema(numeric_range(0, 10))`), and `field()`
definitions.

## Open Data Contract Standard

The [Open Data Contract Standard
v3](https://bitol-io.github.io/open-data-contract-standard/) (ODCS) is
an industry format for describing data products. typethis goes both
ways: typed models export to ODCS YAML and existing ODCS contracts
import back into the model registry.

### Export

```{r, eval = requireNamespace("yaml", quietly = TRUE)}
define_model("Order", fields = list(
  order_id = field("character",
    primary_key = TRUE,
    validator   = string_pattern("^ORD-[0-9]+$")
  ),
  amount = field("numeric",
    validator = numeric_range(0, 1e6),
    pii       = FALSE
  ),
  status = field(t_enum(c("new", "paid", "shipped")),
    default = "new"
  ),
  customer = field("character",
    classification = "confidential",
    pii            = TRUE
  )
))

contract <- to_datacontract("Order",
  info = list(
    name = "orders",
    version = "1.0.0",
    description = "Order records"
  )
)

str(contract, max.level = 2)
```

`write_datacontract()` is the file-IO convenience wrapper:

```{r, eval = requireNamespace("yaml", quietly = TRUE)}
tmp <- tempfile(fileext = ".yaml")
write_datacontract("Order", tmp,
  info = list(name = "orders", version = "1.0.0")
)
readLines(tmp, n = 6)
```

### Import

`from_datacontract()` reads a contract back, calls `define_model()` for
every entry, and assigns generated `new_*()` and `update_*()` constructors
to `envir`.

```{r, eval = requireNamespace("yaml", quietly = TRUE)}
env <- new.env()
from_datacontract(tmp, envir = env)
ls(env)

env$new_Order(order_id = "ORD-1", amount = 42, customer = "Ada")
```

### Field metadata

`field()` accepts ODCS-specific metadata that round-trips through the
contract but has no effect on runtime validation:

| Argument | Used for |
| --- | --- |
| `primary_key` | Marks the field as part of the primary key. |
| `unique` | Indicates uniqueness. |
| `pii` | Personally identifiable information flag. |
| `classification` | `"public"`, `"internal"`, `"confidential"`, ... |
| `tags` | Free-form character vector of tags. |
| `examples` | Example values (also surfaces in JSON Schema). |
| `references` | Foreign-key style reference, e.g. `list(model = "Order", field = "id")`. |
| `quality` | List of engine-specific quality checks. |

### CLI

If the upstream [`datacontract` CLI](https://cli.datacontract.com/) is
installed and on `PATH`, three thin wrappers run it directly from R:

```{r, eval = FALSE}
if (datacontract_cli_available()) {
  datacontract_lint(tmp)
  datacontract_test(tmp, server = "production")
  datacontract_export(tmp, format = "jsonschema")
}
```

## OpenAPI 3.1

OpenAPI 3.1 is JSON-Schema-Draft-2020-12-compatible, so the JSON Schema
machinery powers the OpenAPI bridge as well. Models become entries under
`components.schemas`; `$ref` strings are rewritten from `#/$defs/X` to
`#/components/schemas/X`. Typed functions become a `paths` entry whose
`requestBody` carries the arguments and whose `200` response carries the
return type.

### Models

```{r, eval = requireNamespace("yaml", quietly = TRUE)}
define_model("Address2", fields = list(
  street = field("character"),
  city   = field("character")
))
define_model("PersonDoc", fields = list(
  id = field("integer",
    primary_key = TRUE,
    validator = numeric_range(min = 1L)
  ),
  name = field("character"),
  address = field("Address2")
))

doc <- to_openapi(list("PersonDoc", "Address2"),
  info = list(title = "People API", version = "1.0.0")
)
names(doc$components$schemas)
```

### Typed functions

```{r, eval = requireNamespace("yaml", quietly = TRUE)}
greet <- typed_function(
  function(name, greeting = "Hi") paste(greeting, name),
  arg_specs = list(name = "character", greeting = "character"),
  return_spec = "character"
)
attr(greet, "openapi_op_id") <- "greet"

doc <- to_openapi(list("PersonDoc", greet),
  info = list(title = "People API", version = "1.0.0")
)
names(doc$paths)
```

### Round-trip via disk

```{r, eval = requireNamespace("yaml", quietly = TRUE)}
tmp <- tempfile(fileext = ".yaml")
write_openapi("PersonDoc", tmp,
  info = list(title = "People API", version = "1.0.0")
)

env <- new.env()
from_openapi(tmp, envir = env)
ls(env)
```

`write_openapi()` infers the format from the file extension. Pass
`format = "json"` (or use a `.json` path) for JSON output.

## Choosing a bridge

| If you want to… | Use |
| --- | --- |
| Hand a schema to any JSON-Schema-aware tool | `to_json_schema()` |
| Publish a data product description | `to_datacontract()` |
| Document a JSON HTTP API | `to_openapi()` |
| Bring an existing data contract into R | `from_datacontract()` |
| Bring an existing OpenAPI document into R | `from_openapi()` |
