---
title: "Type-checking your whole package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Type-checking your whole package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

You have an existing R package and want every function inside it to
become typed. Rewriting each function with `typed_function()` would be
a large diff. This vignette shows how to do it in three lines of code,
without touching any existing function definition.

```{r setup}
library(typethis)
```

## The one-line setup

Add a single file `R/zzz.R` to your package:

```r
.onLoad <- function(libname, pkgname) {
  typethis::enable_for_package(pkgname)
}
```

That's it. When R loads your package, `enable_for_package()` walks the
namespace, infers a type spec for every argument with a literal atomic
default, and replaces each binding with a typed wrapper.

## Why this works

`.onLoad()` is called by R *before* the namespace bindings are locked,
so this is the one window where typethis can rewrite each function in
place. Calling the same function from a console or a test (after the
namespace is already locked) will still work, but most bindings will
already be locked and skipped with a single warning.

## What gets typed

Inference covers any argument whose default is a length-1 literal of
a recognised atomic type:

```{r}
infer_specs(function(x = 1L, y = 1.0, name = "a", flag = TRUE) NULL)
```

Arguments without inferable defaults — no default, `NULL` default,
calls like `list()`, `...` — are left unchecked. Their function still
becomes a typed wrapper, just with fewer specs attached.

## Filling in the gaps

Pass `.specs` to add overrides for any function you want to type more
strictly than inference can figure out by itself:

```{r}
ns <- new.env()
ns$add <- function(x = 0L, y = 0L) x + y
ns$normalise <- function(s) tolower(trimws(s))

enable_for_package(ns, .specs = list(
  add = list(.return = "integer"),
  normalise = list(s = "character", .return = "character")
))

attr(ns$add, "return_spec")
attr(ns$normalise, "arg_specs")
```

In your package this would live in `R/zzz.R` next to `.onLoad`:

```r
.onLoad <- function(libname, pkgname) {
  typethis::enable_for_package(
    pkgname,
    .specs = list(
      normalise = list(s = "character", .return = "character"),
      parse_date = list(x = "character", .return = t_predicate(inherits, "Date"))
    )
  )
}
```

## What is skipped

`enable_for_package()` skips two categories automatically:

* **Primitives** — there is nothing to wrap.
* **Package hooks** — `.onLoad`, `.onAttach`, `.onUnload`, `.onDetach`,
  `.Last.lib`, `.First.lib`. These are called by R's namespace
  machinery with a fixed argument shape and should not be wrapped.

You can narrow further with `.filter` (a `function(name, fn)` returning
`TRUE` to retrofit, `FALSE` to skip). The user filter runs *after* the
built-in skip list, so it can only narrow, never widen:

```{r}
ns2 <- new.env()
ns2$public <- function(x = 1L) x
ns2$internal_helper <- function(x = 1L) x

enable_for_package(ns2, .filter = function(name, fn) {
  !startsWith(name, "internal_")
})

is_typed(ns2$public)
is_typed(ns2$internal_helper)
```

## Verifying the result

After your package loads, every typed binding answers `is_typed()`:

```{r}
ns3 <- new.env()
ns3$add <- function(x = 0L, y = 0L) x + y
enable_for_package(ns3)

is_typed(ns3$add)
get_signature(ns3$add)
```

Calling `add()` with the wrong type now produces the typethis error
you'd get from a hand-written `typed_function()`:

```{r, error = TRUE}
ns3$add("a", 3L)
```

## Lifting types from your existing roxygen docs

If your package already has prose documentation, swap
`enable_for_package()` for `as_typed_from_roxygen()`. It reads the
installed `.Rd` files (which roxygen generates), extracts a type spec
for every documented `@param` and `@return`, then forwards to
`enable_for_package()`. Specs derived from docs cover *every*
documented argument, including those without literal defaults — which
is the gap inference cannot fill.

```r
.onLoad <- function(libname, pkgname) {
  typethis::as_typed_from_roxygen(pkgname)
}
```

Two extraction layers run in sequence:

1. **Explicit `[type]` tag.** A `@param` description that begins with
   bracketed text is read verbatim:

   ```r
   #' @param x [integer] Number of iterations
   ```

   Anything between the brackets becomes the spec string — useful when
   prose is ambiguous or the type is one of typethis's compound spec
   constructors (e.g. `[t_vector_of("integer", min_length = 1)]`).
2. **Vocabulary heuristic.** Otherwise the leading prose is matched
   against [`default_type_vocabulary()`]. Patterns are anchored to the
   start of the description, so "A numeric vector" matches but
   "Returns a numeric value when..." does not.

Demonstration with a synthetic Rd directory:

```{r}
rd_dir <- tempfile("rd")
dir.create(rd_dir)
writeLines(c(
  "\\name{add}",
  "\\alias{add}",
  "\\title{Add two numbers}",
  "\\arguments{",
  "\\item{x}{A numeric vector.}",
  "\\item{y}{[integer] Number of times.}",
  "}",
  "\\value{A numeric vector.}"
), file.path(rd_dir, "add.Rd"))

ns4 <- new.env()
ns4$add <- function(x, y) rep(x, y)

as_typed_from_roxygen(ns4, .rd_dir = rd_dir)

attr(ns4$add, "arg_specs")
attr(ns4$add, "return_spec")
```

Inference still runs alongside, so undocumented arguments with literal
defaults still get specs from `infer_specs()`. User `.specs` win over
both, just like with `enable_for_package()`.

To extend the prose vocabulary — e.g. recognise `tbl_df` as
`data.frame` — prepend an entry:

```r
my_vocab <- c(
  "(?i)^\\s*(?:a|an)?\\s*tbl[._]df\\b" = "data.frame",
  default_type_vocabulary()
)
as_typed_from_roxygen("mypkg", .vocabulary = my_vocab)
```

## Trade-offs

* **Coverage scales with how much you've used defaults.** With
  `enable_for_package()` alone, functions written without literal
  defaults (e.g. `function(x, y) ...`) get a typed wrapper but no
  specs unless you fill them in via `.specs`.
  `as_typed_from_roxygen()` plugs that gap by lifting types out of
  your existing prose.
* **Heuristics are imperfect.** The vocabulary recognises common type
  language; ambiguous prose silently produces no spec. Use the
  `[type]` prefix or `.specs` overrides for anything you care about
  precisely.
* **Per-call overhead.** Every call now passes through a wrapper that
  validates arguments. For hot paths, set `.validate = FALSE` per call
  via `validate_call()` or pass `.validate = FALSE` to
  `enable_for_package()` / `as_typed_from_roxygen()`.
* **CRAN.** Modifying your own namespace from your own `.onLoad()` is
  fine. Modifying *another* package's namespace from outside is a
  separate, more invasive pattern that these functions deliberately
  do not support.

## Related entry points

* [`as_typed()`][as_typed] — retrofit one function at a time.
* [`as_typed_env()`][as_typed_env] — the engine `enable_for_package()`
  is built on; gives you full control over filtering and overrides.
* [`as_typed_from_roxygen()`][as_typed_from_roxygen] — lift types out
  of prose `@param`/`@return` documentation.
* [`parse_param_type()`][parse_param_type] — inspect what would be
  derived from a single description string.
* [`typed_function()`][typed_function] — write a typed function from
  scratch instead of retrofitting.
