---
title: "Getting Started with typethis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with typethis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = TRUE
)
```

# Introduction

`typethis` brings comprehensive type safety and validation to R, inspired by Python's `mypy` and `pydantic`. This guide will walk you through all the major features.

```{r setup}
library(typethis)
```

## Basic Type Checking

The foundation of `typethis` is runtime type checking. You can check, assert, and validate types easily.

### Checking Types

```{r}
# Basic type checking
is_type(5, "numeric")
is_type("hello", "character")
is_type(TRUE, "logical")

# Check with NULL handling
is_type(NULL, "numeric") # FALSE
is_type(NULL, "numeric", nullable = TRUE) # TRUE

# Check multiple possible types
is_one_of(5, c("numeric", "character"))
is_one_of("hello", c("numeric", "character"))
```

### Asserting Types

Use `assert_type()` when you want to enforce type constraints and throw errors on mismatch:

```{r}
# Successful assertion
assert_type(5, "numeric", "my_variable")

# Failed assertion (will throw error)
# assert_type("hello", "numeric", "my_variable")
```

### Validating Types

For more controlled error handling, use `validate_type()`:

```{r}
# Successful validation
result <- validate_type(5, "numeric", "x")
result$valid
result$error

# Failed validation
result <- validate_type("hello", "numeric", "x")
result$valid
result$error
```

## Type Coercion

Safely convert between types with validation:

```{r}
# Basic coercion
coerce_type("123", "numeric")
coerce_type(123, "character")
coerce_type(c(1, 2, 3), "integer")

# Strict mode (fails on NA coercion)
# coerce_type("abc", "numeric", strict = TRUE)  # Error
```

## Typed Functions

Create functions with automatic type checking for arguments and return values.

### Basic Typed Functions

```{r}
# Define a typed function
add_numbers <- typed_function(
  fn = function(x, y) x + y,
  arg_types = list(x = "numeric", y = "numeric"),
  return_type = "numeric"
)

# Valid call
add_numbers(5, 3)

# Invalid call (will error)
# add_numbers("a", "b")
```

### Calling Conventions (v0.2+)

`typed_function()` supports all R calling conventions:

```{r}
add <- typed_function(
  fn = function(x, y) x + y,
  arg_specs = c(x = "numeric", y = "numeric")
)

# Positional arguments
add(1, 2)

# Named arguments
add(x = 1, y = 2)

# Reordered named arguments
add(y = 2, x = 1)

# Mixed positional and named
add(1, y = 2)
```

### Missing Argument Detection (v0.2+)

```{r}
add <- typed_function(
  fn = function(x, y) x + y,
  arg_specs = c(x = "numeric", y = "numeric")
)

# Missing required argument
# add(1)  # Error: missing required argument 'y' with no default

# Arguments with defaults are optional
add_with_default <- typed_function(
  fn = function(x, y = 1) x + y,
  arg_specs = c(x = "numeric", y = "numeric")
)
add_with_default(5) # Returns 6
```

### Variadic Functions (v0.2+)

```{r}
# ... passthrough works correctly
sum_fn <- typed_function(
  fn = function(x, ...) sum(x, ...),
  arg_specs = c(x = "numeric")
)
sum_fn(c(1, NA, 3), na.rm = TRUE)
```

### Function Signatures

Use signatures for cleaner code:

```{r}
# Define a signature
sig <- signature(
  x = "numeric",
  y = "numeric",
  .return = "numeric"
)

# Apply signature to function
multiply <- with_signature(function(x, y) x * y, sig)
multiply(5, 3)

# Check if function is typed
is_typed(multiply)

# Get function signature (v0.2+ includes formals)
get_signature(multiply)
```

### Type Coercion in Functions

```{r}
# Function with automatic type coercion
add_flexible <- typed_function(
  fn = function(x, y) x + y,
  arg_types = list(x = "numeric", y = "numeric"),
  return_type = "numeric",
  coerce = TRUE
)

# This works because strings are coerced to numeric
add_flexible("5", "3")
```

### Validating Calls

Validate function calls without executing:

```{r}
result <- validate_call(add_numbers, x = 5, y = 3)
result$valid

result <- validate_call(add_numbers, x = "a", y = 3)
result$valid
result$errors
```

## Advanced Validators

`typethis` provides many built-in validators for common validation patterns.

### Numeric Range Validator

```{r}
# Age validator (0-120)
age_validator <- numeric_range(min = 0, max = 120)
age_validator(25)
age_validator(150)

# Percentage validator (0-100, exclusive bounds)
percent_validator <- numeric_range(
  min = 0,
  max = 100,
  exclusive_min = FALSE,
  exclusive_max = FALSE
)
percent_validator(50)
```

### String Validators

```{r}
# String length validator
name_validator <- string_length(min_length = 1, max_length = 50)
name_validator("John Doe")
name_validator("")

# Email pattern validator
email_validator <- string_pattern(
  "^[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}$"
)
email_validator("user@example.com")
email_validator("invalid-email")
```

### Collection Validators

```{r}
# Vector length
pair_validator <- vector_length(exact_len = 2)
pair_validator(c(1, 2))
pair_validator(c(1, 2, 3))

# List of specific type
num_list_validator <- list_of("numeric", min_length = 1, max_length = 5)
num_list_validator(list(1, 2, 3))
num_list_validator(list("a", "b"))
```

### Data Frame Validators

```{r}
# Data frame with required columns
df_validator <- dataframe_spec(
  required_cols = c("id", "name", "age"),
  min_rows = 1,
  max_rows = 1000
)

df_valid <- data.frame(
  id = 1:3,
  name = c("Alice", "Bob", "Charlie"),
  age = c(25, 30, 35)
)
df_validator(df_valid)

df_invalid <- data.frame(id = 1:3, name = c("Alice", "Bob", "Charlie"))
df_validator(df_invalid)
```

### Enum Validators

```{r}
# Status enum
status_validator <- enum_validator(c("active", "inactive", "pending"))
status_validator("active")
status_validator("deleted")
```

### Combining Validators

```{r}
# Positive number validator
positive_num <- combine_validators(
  function(x) is.numeric(x),
  function(x) all(x > 0),
  all_of = TRUE
)
positive_num(5)
positive_num(-5)

# Number OR string validator
num_or_str <- combine_validators(
  function(x) is.numeric(x),
  function(x) is.character(x),
  all_of = FALSE
)
num_or_str(5)
num_or_str("hello")
num_or_str(TRUE)
```

### Nullable Validators

```{r}
# Optional numeric (can be NULL)
optional_num <- nullable(function(x) is.numeric(x))
optional_num(5)
optional_num(NULL)
optional_num("hello")
```

## Composable Type Specs (v0.3+)

Beyond plain type names like `"numeric"`, `typethis` provides a small set of constructors that compose into richer specifications. They work everywhere a type name does — `is_type()`, `assert_type()`, `typed_function()`, and `field()`.

```{r}
# Either an integer or a character
spec <- t_union("integer", "character")
is_type(1L, spec)
is_type("x", spec)
is_type(1.5, spec)

# A value that may also be NULL
maybe_num <- t_nullable("numeric")
is_type(NULL, maybe_num)
is_type(3.14, maybe_num)

# A list whose elements are all integers, or a vector of them
is_type(list(1L, 2L, 3L), t_list_of("integer"))
is_type(c(1L, 2L, 3L), t_vector_of("integer"))

# A finite enumeration
status <- t_enum(c("active", "inactive", "pending"))
is_type("active", status)
is_type("deleted", status)
```

Specs compose freely. The example below accepts a list whose elements are each either an integer or a character:

```{r}
mixed <- t_list_of(t_union("integer", "character"))
is_type(list(1L, "two", 3L), mixed)
```

Use `is_type_spec()` to detect a composite spec, and pass any spec to `field()` when defining a model.

## Typed Models

The most powerful feature of `typethis` - create typed data models with automatic validation.

### Basic Models

```{r}
# v0.2+ New-style API: define_model("ClassName", fields = list(...))
# This creates new_ClassName() and update_ClassName() in your environment
define_model("User",
  fields = list(
    name = field("character", nullable = FALSE),
    age = field("numeric", nullable = FALSE),
    email = field("character", nullable = FALSE)
  )
)

# Create a valid user
user <- new_User(
  name = "John Doe",
  age = 30,
  email = "john@example.com"
)

# Access fields
user$name
user$age

# Check if it's a model
is_model(user)

# Old-style API (still supported for backward compatibility)
Person <- define_model(
  name = "character",
  age = "numeric",
  email = "character"
)
person <- Person(name = "Jane", age = 25, email = "jane@example.com")
```

### Fields with Defaults

```{r}
# v0.2+ New-style API with defaults
define_model("Config",
  fields = list(
    host = field("character", default = "localhost"),
    port = field("integer", default = 8080L),
    debug = field("logical", default = FALSE)
  )
)

# Create with defaults
config1 <- new_Config()
config1$host
config1$port

# Override defaults
config2 <- new_Config(host = "example.com", port = 3000L)
config2$host
```

### Fields with Custom Validators

```{r}
# v0.2+ New-style API with field validators
define_model("Person",
  fields = list(
    name = field(
      type = "character",
      validator = string_length(min_length = 1, max_length = 100),
      description = "Person's full name"
    ),
    age = field(
      type = "numeric",
      validator = numeric_range(min = 0, max = 120),
      default = 0,
      description = "Person's age in years"
    ),
    email = field(
      type = "character",
      validator = string_pattern(
        "^[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}$"
      ),
      description = "Email address"
    )
  )
)

# Valid person
person <- new_Person(
  name = "Jane Doe",
  age = 28,
  email = "jane@example.com"
)

# Invalid age (will error)
# new_Person(name = "John", age = 150, email = "john@example.com")

# Invalid email (will error)
# new_Person(name = "John", age = 30, email = "invalid-email")
```

### Strict Mode

```{r}
# v0.2+ Strict mode: no extra fields allowed
define_model("StrictUser",
  fields = list(
    name = field("character"),
    age = field("numeric")
  ),
  .strict = TRUE
)

# This works
user <- new_StrictUser(name = "John", age = 30)

# This fails (extra field 'email')
# new_StrictUser(name = "John", age = 30, email = "john@example.com")
```

### Working with Models

```{r}
# Get model schema
schema <- get_schema(User)
names(schema)

# v0.2+ Update model fields with update_<ClassName>()
person <- new_Person(name = "Jane", age = 28, email = "jane@example.com")
person <- update_Person(person, age = 29)
person$age

# Validate model instance
result <- validate_model(person)
result$valid

# Convert to list
person_list <- model_to_list(person)
person_list
```

## JSON Schema Export (v0.3+)

`to_json_schema()` serializes typed models, type specs, and validators into a JSON Schema (Draft 2020-12) fragment. Builtin validator factories (`numeric_range()`, `string_length()`, `string_pattern()`, `enum_validator()`, ...) attach a `constraint` attribute to their closure that the exporter reads — so range and pattern constraints surface as native `minimum` / `maxLength` / `pattern` keys.

```{r json-schema, eval = requireNamespace("jsonlite", quietly = TRUE)}
define_model("Account",
  fields = list(
    id      = field("integer", validator = numeric_range(min = 1L)),
    email   = field("character", validator = string_pattern("^[^@]+@[^@]+$")),
    role    = field("character",
                    validator = enum_validator(c("admin", "user"))),
    tags    = field(t_list_of("character"), default = list())
  )
)

schema <- to_json_schema("Account")
jsonlite::toJSON(schema, auto_unbox = TRUE, pretty = TRUE)
```

Constructs without a canonical JSON Schema mapping (data frames, factors, custom predicate functions) are emitted as `x-typethis-*` extension keys so that the schema round-trips through typethis-aware tooling without losing information. Nested models become `$ref` entries with auto-populated `$defs`; cyclic references are handled via a stub-then-fill protocol.

## Data Contracts (ODCS v3, v0.4+)

The `to_datacontract()` / `from_datacontract()` bridge maps typed models to and from the [Open Data Contract Standard v3](https://github.com/bitol-io/open-data-contract-standard). `field()` accepts ODCS metadata arguments — `primary_key`, `unique`, `pii`, `classification`, `tags`, `examples`, `references`, `quality` — that round-trip through the contract but have no effect on runtime validation.

```{r datacontract, eval = requireNamespace("yaml", quietly = TRUE)}
define_model("Customer",
  fields = list(
    customer_id = field("integer",
      primary_key = TRUE,
      unique = TRUE,
      description = "Surrogate key"
    ),
    email = field("character",
      pii = TRUE,
      classification = "confidential",
      validator = string_pattern("^[^@]+@[^@]+$")
    ),
    signup_date = field("date",
      tags = c("audit", "lifecycle")
    )
  )
)

contract <- to_datacontract("Customer",
  info = list(name = "Customers", version = "1.0.0")
)
str(contract, max.level = 2)
```

Write the contract to disk with `write_datacontract()` and read it back with `from_datacontract()` (which also re-registers the schemas as `new_*()` constructors in the calling environment):

```{r datacontract-roundtrip, eval = requireNamespace("yaml", quietly = TRUE)}
tmp <- tempfile(fileext = ".yaml")
write_datacontract("Customer", tmp,
  info = list(name = "Customers", version = "1.0.0")
)

env <- new.env()
from_datacontract(tmp, envir = env)
ls(env)
```

For full validation against the standard, the upstream [`datacontract` CLI](https://datacontract.com) is available through thin wrappers — `datacontract_lint()`, `datacontract_test()`, `datacontract_export()` — gated by `datacontract_cli_available()`:

```{r datacontract-cli, eval = FALSE}
if (datacontract_cli_available()) {
  datacontract_lint(tmp)
}
```

## OpenAPI 3.1 Export (v0.5+)

OpenAPI 3.1 is JSON-Schema-Draft-2020-12-compatible, so the same machinery that backs `to_json_schema()` also drives `to_openapi()`. Models become entries under `components.schemas`; nested-model `$ref` strings are rewritten from JSON Schema's `#/$defs/X` form to OpenAPI's `#/components/schemas/X` form. Typed functions become a single `paths` entry whose `requestBody` carries the arguments as a JSON object and whose `200` response carries the return type.

```{r openapi-models, eval = requireNamespace("yaml", quietly = TRUE)}
define_model("Address", fields = list(
  street = field("character"),
  city   = field("character")
))
define_model("Person", fields = list(
  id      = field("integer", primary_key = TRUE,
                  validator = numeric_range(min = 1L)),
  name    = field("character"),
  address = field("Address")
))

doc <- to_openapi(list("Person", "Address"),
                  info = list(title = "People API", version = "1.0.0"))
str(doc, max.level = 3)
```

Typed functions can be added to the same document. Set `attr(fn, "openapi_op_id")` to control the operation id (and the path); otherwise the default is `/operation`.

```{r openapi-function, eval = requireNamespace("yaml", quietly = TRUE)}
greet <- typed_function(
  function(name, greeting = "Hi") paste(greeting, name),
  arg_specs = list(name = "character", greeting = "character"),
  return_spec = "character"
)
attr(greet, "openapi_op_id") <- "greet"

doc <- to_openapi(list("Person", greet),
                  info = list(title = "People API", version = "1.0.0"))
names(doc$paths)
```

Round-trip through disk works the same way as the Data Contract bridge:

```{r openapi-roundtrip, eval = requireNamespace("yaml", quietly = TRUE)}
tmp <- tempfile(fileext = ".yaml")
write_openapi("Person", tmp,
              info = list(title = "People API", version = "1.0.0"))

env <- new.env()
from_openapi(tmp, envir = env)
ls(env)
```

`write_openapi()` infers the format from the file extension — pass `format = "json"` (or use a `.json` path) for JSON output, `format = "yaml"` for YAML.

## Real-World Examples

### API Request Validation

```{r}
# Define API request model
APIRequest <- define_model(
  method = field(
    type = "character",
    validator = enum_validator(c("GET", "POST", "PUT", "DELETE"))
  ),
  endpoint = field(
    type = "character",
    validator = string_pattern("^/api/")
  ),
  headers = field(
    type = "list",
    default = list()
  ),
  body = field(
    type = "list",
    default = list(),
    nullable = TRUE
  ),
  .strict = TRUE
)

# Create valid request
request <- APIRequest(
  method = "GET",
  endpoint = "/api/users",
  headers = list("Authorization" = "Bearer token123")
)
```

### Data Processing Pipeline

```{r}
# Typed data processing function
process_sales <- typed_function(
  fn = function(data, min_amount) {
    data[data$amount >= min_amount, ]
  },
  arg_types = list(
    data = dataframe_spec(required_cols = c("id", "amount", "date")),
    min_amount = numeric_range(min = 0)
  ),
  return_type = "data.frame"
)

# Example usage
sales_data <- data.frame(
  id = 1:5,
  amount = c(100, 250, 50, 300, 150),
  date = as.Date("2024-01-01") + 0:4
)

# Process with validation
filtered <- process_sales(sales_data, min_amount = 100)
filtered
```

### Configuration Management

```{r}
# Database configuration model
DatabaseConfig <- define_model(
  host = field("character", default = "localhost"),
  port = field(
    "integer",
    validator = numeric_range(min = 1, max = 65535),
    default = 5432L
  ),
  database = "character",
  user = "character",
  password = "character",
  ssl = field("logical", default = FALSE),
  .strict = TRUE
)

# Application configuration
AppConfig <- define_model(
  app_name = "character",
  version = field("character", default = "1.0.0"),
  debug = field("logical", default = FALSE),
  max_connections = field(
    "integer",
    validator = numeric_range(min = 1, max = 1000),
    default = 10L
  )
)

# Create configuration
db_config <- DatabaseConfig(
  database = "myapp",
  user = "admin",
  password = "secret123"
)

app_config <- AppConfig(
  app_name = "My Application",
  debug = TRUE
)
```

## Best Practices

### 1. Use Type Checking at Boundaries

```{r}
# Good: validate external inputs
process_user_input <- function(user_data) {
  assert_type(user_data, "list", "user_data")

  # Process data...
  user_data
}
```

### 2. Combine with Existing Code

```{r}
# Wrap existing functions with type safety
unsafe_add <- function(x, y) x + y

safe_add <- typed_function(
  fn = unsafe_add,
  arg_types = list(x = "numeric", y = "numeric"),
  return_type = "numeric"
)
```

### 3. Use Models for Complex Data Structures

```{r}
# Instead of unnamed lists, use models
# Bad:
# user <- list(name = "John", age = 30)

# Good:
User <- define_model(name = "character", age = "numeric")
user <- User(name = "John", age = 30)
```

### 4. Leverage Custom Validators

```{r}
# Create reusable validators
is_positive <- function(x) is.numeric(x) && all(x > 0)
is_non_empty <- function(x) is.character(x) && all(nchar(x) > 0)

Product <- define_model(
  name = field("character", validator = is_non_empty),
  price = field("numeric", validator = is_positive),
  quantity = field("integer", validator = is_positive)
)
```

## Performance Considerations

Type checking adds runtime overhead. Here are some tips:

1. **Validate at boundaries**: Check types when data enters your system
2. **Disable validation in production**: Use `.validate = FALSE` for performance-critical code
3. **Use lazy validation**: Validate only when needed
4. **Cache validators**: Reuse validator functions

```{r}
# Disable validation for performance
define_model("FastUser",
  fields = list(
    name = field("character"),
    age = field("numeric")
  ),
  .validate = FALSE # Skip validation
)

# Or validate selectively
user <- new_FastUser(name = "John", age = 30)
validate_model(user) # Validate later if needed
```

## Runtime Only

**Important**: `typethis` provides **runtime validation only**, not static type checking.

- Type errors are caught when R code executes, not during development
- It does not replace static analysis tools like `mypy` for Python
- Benefits: works with any R code, no special IDE or tooling required
- Trade-off: errors are only caught at runtime

If you need static analysis for R, consider tools like `lintr` or IDE-based diagnostics.

## Summary

`typethis` provides comprehensive type safety for R:

- ✅ Runtime type checking with `is_type()`, `assert_type()`, `validate_type()`
- ✅ Typed functions with `typed_function()` and `signature()`
- ✅ Advanced validators for common patterns
- ✅ Typed models similar to Pydantic
- ✅ Type coercion with validation
- ✅ Easy integration with existing code

Start using `typethis` today to write more robust, maintainable R code!
