---
title: "Composable type specifications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Composable type specifications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(typethis)
```

A *type spec* is a structured, composable type description. The `t_*()`
constructors build them; they work everywhere a plain type name does —
`is_type()`, `assert_type()`, `validate_type()`, `field()`,
`typed_function()`, and `to_json_schema()`.

Plain character builtins (`"numeric"`) and predicate functions still work
as type arguments without change. Type specs are an additive layer for
when those aren't expressive enough.

## The constructors

| Constructor | Matches |
| --- | --- |
| `t_union(...)` | any of the alternatives |
| `t_nullable(type)` | `type` or `NULL` |
| `t_list_of(type, …)` | a list whose every element matches `type` |
| `t_vector_of(type, …)` | an atomic vector of a builtin type |
| `t_enum(values)` | values in a fixed allowed set |
| `t_model(class)` | an instance of a registered typed model |
| `t_predicate(fn, description)` | predicate with a documented description |

Use `is_type_spec()` to detect a composite spec at runtime.

## Examples

### Union

```{r}
id <- t_union("integer", "character")
is_type(1L, id)
is_type("u-1", id)
is_type(1.5, id)
```

### Nullable

```{r}
maybe_int <- t_nullable("integer")
is_type(NULL, maybe_int)
is_type(1L, maybe_int)
is_type("hi", maybe_int)
```

### Lists and vectors

```{r}
tags <- t_list_of("character", min_length = 1L)
is_type(list("alpha", "beta"), tags)
is_type(list(), tags)

triple <- t_vector_of("integer", exact_length = 3L)
is_type(1:3, triple)
is_type(1:5, triple)
```

`t_list_of()` accepts any element type spec, including other composites:

```{r}
mixed <- t_list_of(t_union("integer", "character"))
is_type(list(1L, "two", 3L), mixed)
```

### Enum

```{r}
role <- t_enum(c("admin", "user", "guest"))
is_type("admin", role)
is_type("root", role)
```

`t_enum()` is the type-spec sibling of [`enum_validator()`](
../reference/enum_validator.html). Use the type-spec form when you want
the enum to appear as a true type in `field()` and `typed_function()` and
to surface in JSON Schema as a typed `enum` keyword.

### Model references

References resolve at validation time, which makes forward references
(and cycles) possible:

```{r}
define_model("Address", fields = list(zip = field("character")))
addr <- t_model("Address")
is_type(new_Address(zip = "10115"), addr)
```

### Predicates with a description

`t_predicate()` is a thin wrapper around a custom predicate that carries
a description. The description surfaces in error messages and in JSON
Schema output.

```{r}
positive <- t_predicate(
  function(x) is.numeric(x) && all(x > 0),
  description = "positive number"
)
is_type(5, positive)
is_type(-1, positive)
```

## Using specs in models and functions

Anywhere `type =` or `arg_specs =` accepts a builtin name, it also
accepts a spec.

```{r}
define_model("Order", fields = list(
  id     = field(t_union("integer", "character")),
  status = field(t_enum(c("new", "paid", "shipped")), default = "new"),
  tags   = field(t_list_of("character"), default = list())
))

ord <- new_Order(id = "ord-1", tags = list("priority", "rush"))
ord$status
ord$tags
```

```{r}
greet <- typed_function(
  function(name, n = 1L) paste(rep(name, n), collapse = " "),
  arg_specs = list(
    name = "character",
    n    = t_union("integer", "numeric")
  ),
  return_spec = "character"
)
greet("hi", n = 3)
```

## Coercion

`coerce_type()` understands a subset of specs where coercion has a clear
meaning:

- `t_nullable(...)` — `NULL` passes through; otherwise the inner spec
  drives coercion;
- `t_union(...)` — each alternative is tried in order; the first that
  coerces and validates wins;
- `t_enum(...)` — values already in the allowed set pass through;
  otherwise the value is coerced to the enum's value type and re-checked.

Other kinds (`t_list_of()`, `t_vector_of()`, `t_model()`, `t_predicate()`)
raise an explicit error.

```{r}
coerce_type("paid", t_enum(c("new", "paid", "shipped")))
coerce_type(NULL, t_nullable("integer"))
coerce_type("123", t_union("integer", "character"))
```
