---
title: "Validators and models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validators and models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(typethis)
```

This vignette covers the built-in validators and the full `define_model()`
/ `field()` API. If you haven't read [Getting
Started](getting-started.html) yet, start there.

## Validators

A validator is just a function `function(value) -> logical`. The package
ships with factories for the most common rules; their results carry a
`constraint` attribute so that bridges (JSON Schema, ODCS, OpenAPI) can
emit native keywords (`minimum`, `maxLength`, `pattern`, ...) instead of
opaque predicate stubs.

### Numeric ranges

```{r}
age <- numeric_range(0, 120)
age(25)
age(150)

# Probability strictly in (0, 1)
prob <- numeric_range(0, 1, exclusive_min = TRUE, exclusive_max = TRUE)
prob(0.5)
prob(0)
```

### Strings

```{r}
name <- string_length(min_length = 1, max_length = 50)
name("Ada Lovelace")
name("")

email <- string_pattern("^[^@]+@[^@]+\\.[^@]+$")
email("user@example.com")
email("not-an-email")
```

### Vectors and lists

```{r}
pair <- vector_length(exact_len = 2)
pair(c(1, 2))
pair(c(1, 2, 3))

nums <- list_of("numeric", min_length = 1)
nums(list(1, 2, 3))
nums(list("a"))

# Optional values: wrap any validator with `nullable()`
optional_num <- nullable(function(x) is.numeric(x))
optional_num(NULL)
optional_num(5)
```

### Data frames

```{r}
is_orders <- dataframe_spec(
  required_cols = c("id", "amount"),
  min_rows = 1
)
is_orders(data.frame(id = 1:3, amount = c(10, 20, 30)))
is_orders(data.frame(id = integer()))
is_orders(data.frame(name = "Ada"))
```

### Enums

```{r}
status <- enum_validator(c("active", "inactive", "pending"))
status("active")
status("deleted")
```

For the equivalent that doubles as a [type
spec](type-specs.html) — usable inside `field()` and `typed_function()` —
see `t_enum()`.

### Combining validators

```{r}
positive_num <- combine_validators(
  function(x) is.numeric(x),
  function(x) all(x > 0)
)
positive_num(5)
positive_num(-1)

num_or_str <- combine_validators(
  function(x) is.numeric(x),
  function(x) is.character(x),
  all_of = FALSE
)
num_or_str(5)
num_or_str("hi")
num_or_str(TRUE)
```

### Reading constraint metadata

`validator_constraint()` exposes the structured constraint attached by
the built-in factories — useful for tooling and for the bridge functions:

```{r}
validator_constraint(numeric_range(0, 10))
validator_constraint(function(x) x > 0) # NULL for plain user predicates
```

## Models

Use `define_model()` to describe a record. It creates a `new_<Class>()`
constructor and an `update_<Class>()` helper in the calling environment.

```{r}
define_model("Person", fields = list(
  name = field("character"),
  age = field("integer", validator = numeric_range(0, 120)),
  email = field("character",
    validator = string_pattern("^[^@]+@[^@]+$")
  )
))

p <- new_Person(name = "Ada", age = 36L, email = "ada@example.com")
p
```

### Defaults

```{r}
define_model("Config", fields = list(
  host  = field("character", default = "localhost"),
  port  = field("integer", default = 8080L),
  debug = field("logical", default = FALSE)
))

new_Config()$host
new_Config(host = "api.example.com", port = 443L)$port
```

### Nullable fields

A field with `nullable = TRUE` is allowed to be `NULL` (and to be
omitted at construction without a default). Without it, omission is an
error unless a default is set.

```{r}
define_model("Profile", fields = list(
  name = field("character"),
  bio  = field("character", nullable = TRUE)
))

new_Profile(name = "Ada")$bio
new_Profile(name = "Ada", bio = "Mathematician")$bio
```

### Strict mode

`.strict = TRUE` rejects unknown fields at construction:

```{r, error = TRUE}
define_model("StrictPoint", fields = list(
  x = field("numeric"),
  y = field("numeric")
), .strict = TRUE)

new_StrictPoint(x = 1, y = 2, z = 3)
```

### Nested models

A field's `type` may be a registered model class name. The constructor
will check that the value is an instance of that class.

```{r}
define_model("Address", fields = list(
  street = field("character"),
  city   = field("character")
))

define_model("Person2", fields = list(
  name    = field("character"),
  address = field("Address")
))

new_Person2(
  name    = "Ada",
  address = new_Address(street = "Newstead", city = "Nottingham")
)
```

Use `t_model("Address")` if you want a [type spec](type-specs.html) form
that can be combined with `t_nullable()` or `t_list_of()`.

### Field-level validators

`field()` accepts a `validator` callback that runs after the type check:

```{r, error = TRUE}
define_model("Coupon", fields = list(
  code = field("character",
    validator = string_length(min_length = 4, max_length = 12)
  ),
  discount = field("numeric",
    validator = numeric_range(0, 1, exclusive_max = TRUE)
  )
))

new_Coupon(code = "SAVE20", discount = 0.2)
new_Coupon(code = "AB", discount = 0.2)
```

### Working with instances

```{r}
schema_keys <- names(get_schema(p))
schema_keys

is_model(p)
model_to_list(p)

result <- validate_model(p)
result$valid
```

`update_<Class>()` is the preferred way to mutate — it preserves the S3
class and revalidates:

```{r}
p2 <- update_Person(p, age = 37L)
p2$age
```

For mutation across models without a class-specific updater, use
`update_model()`.

### Field metadata for bridges

`field()` accepts metadata that travels through to the `to_json_schema()`,
`to_datacontract()`, and `to_openapi()` exporters but has no effect on
runtime validation. See the [interop](interop.html) vignette for what
each key does.

```{r}
field(
  "character",
  primary_key    = TRUE,
  pii            = TRUE,
  classification = "confidential",
  tags           = c("audit", "lifecycle"),
  examples       = list("ORD-1", "ORD-2")
)
```

## Performance

Validation has a per-call cost. For hot paths:

- pass `.validate = FALSE` to `define_model()` to disable checks at
  construction;
- pass `validate = FALSE` to `typed_function()` for the same effect on
  function calls;
- check upstream once and trust your boundaries.

You can still call `validate_model()` on demand to re-check an instance
after a series of unchecked updates.
